---
title: "imputation_stickle"
author: "Akhil Ghosh"
output: html_document
---

```{r setup, include=FALSE}
library(mixgb)
library(mice)
library(tidyverse)
library(janitor)
library(missMethods)
library(ROCR)


```



DATA DICTIONARY
stl: standard length. A measure of size from tip of snout to end of tail. (mm)

ips: index pelvic score. a categorical measure of pelvic complexity based on the number of pelvic elements observed. 0-3 in fossils. 0-8 in extant fish.
ips.prop: ips/3 (fossils) or ips/8 (extant) to put ips on same scale from 0 to 1.
lps: length of a pelvic spine (if ips <3 (fossils) or ips < 8 (extant), then pelvic spines were considered evolutionarily lost, and were assigned lengths of 0)
tpg: pelvic girdle length (mm). (when missing (ips = 0), considered evolutionarily lost and assigned lengths of 0.)

mav: abdominal vertebrae number (count)
mcv: caudal vertebrae number (count)
mdf: dorsal fin ray number. number of bones in the dorsal fin (count)
maf: anal fin ray number (count)

mds: dorsal spine number (count)
ds1: length of the first dorsal spine (mm) (when missing, the first dorsal spine was considered evolutionarily lost, and assigned lengths of 0. when spines were broken assigned NA)
ds2: length of the second dorsal spine (mm) (when missing, the second dorsal spine was considered evolutionarily lost, and assigned lengths of 0. when spines were broken assigned NA)
ds3: length of the third dorsal spine (mm) (when missing, the third dorsal spine was considered evolutionarily lost, and assigned lengths of 0. when spines were broken assigned NA)

mpt: pterygiophore number (count). Number of armor bones on the "back". 
lpt: length of the pterygiophore anterior to the the third spine (mm)

ect: ectocoracoid length (mm)
cle: cleithrum length (mm)
pmx: premaxila length (mm)


```{r}
stickle <- read_csv('stickle_revamped.csv')

stick <- stickle %>% 
  mutate(gender = as.factor(case_when(
    grepl(".M.",univID,fixed=TRUE) ~ "Male",
    grepl(".F.",univID,fixed=TRUE) ~ "Female",
    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
    )), ips = as.factor(ips)
  ) %>% 
  select(gender, everything())

stick_whole <- stick %>% select(-c(univID, ips))

extant_stickle <- stick %>% filter(gender != "Fossil") %>% select(!univID)
extant_stickle
summary(extant_stickle)
```



```{r}

#Checking imputation accuracy/viability

```







```{r}
#Using MICE imputation to get imputed datasets
M = 5
stick_whole <- stick %>% select(-c(univID, ips))
summary(stick_whole)
imputed_stick_whole <- mice(stick_whole,m=5)
complete_stick_list <- list()
for(m in 1:M){
  complete_stick_list[[m]] <- complete(imputed_stick_whole, m)
}
```


```{r}
#getting time vector, adding it to imputed data
univ <- stick %>% 
  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
times <- univ %>% drop_na(time) %>% cut(time, breaks = 1,3,5,7,9,11,13,15,17) pull(time)

cut(times, breaks = 8, labels = c(8,7,6,5,4,3,2,1))
times_vector <-length(rep(c(1,2,3,4,5,6,7,8,9),each=86))
cut(times_vector, breaks=2, labels)


help(cut)
```


```{r}
#Initial grouping and summarizing
# temp_comp <- complete_stick_list[[1]][-c(1:367),]
# temp_comp$time <- times_vector
# 
# means_stl_time <- temp_comp %>% group_by(gender, time) %>% summarize(xbar = mean(stl), se = sqrt(var(stl)/n()), n())
# 
# 
# help(mice)
# temp_comp %>% select(gender, time, stl) %>% filter(time==17)
# 
# 
# with(temp_)
# 
# means_stl_time %>% ggplot(aes(x = time, y = xbar, col = gender))+ geom_point() +geom_line()
# 
# table(complete_stick_list[[1]][c(1:367),]$gender)
# 
# table(complete_stick_list[[1]][-c(1:367),]$gender)

```


```{r}
#doing the grouping and summarizing for all imputed datasets, looking specifically at mean standard length for each group

temp_comp_lists <- list()
est_list <- list()
M = 5
for(m in 1:M){

temp_comp_lists[[m]] <- complete_stick_list[[m]][-c(1:367),]
temp_comp_lists[[m]]$time <- times_vector


est_list[[m]] <- temp_comp_lists[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(stl), se = sqrt(var(stl)/n()), n())
est_list[[m]]$imp = m
}


# stacking the datasets
combined_est <- do.call(rbind,est_list)
combined_est
```


```{r}
#Calculating error bars of combined estimates

combined_est_errors <- combined_est %>% 
  group_by(gender, time) %>% 
  summarize(x_bar_bar  = mean(xbar), 
            B_m = var(xbar), 
            W_m = mean(se^2), 
            se_xbar_bar = sqrt(W_m+(1+1/5)*(B_m)), 
            r = (1+1/5)*B_m/W_m, 
            U = (5-1)*(1+(1/r)^2), 
            t_val = qt(0.05/2, df = U, lower.tail = FALSE), 
            uci = x_bar_bar + t_val * se_xbar_bar, 
            lci = x_bar_bar - t_val * se_xbar_bar, 
            t_se = t_val * se_xbar_bar)

combined_est_errors %>% select(gender, time,x_bar_bar, lci, uci)

```


```{r}
#ggplot of the data for 

ggplot(data= combined_est_errors, aes(x = time, y = x_bar_bar, col = gender))+geom_point()+geom_line() + geom_errorbar(aes(ymin = lci, ymax = uci, group = gender),
     width = 0.2)+ labs(x = "Time", y = "Mean of imputed mean standard length values (STL)")



```






```{r}
#Doing it for cle variable this time
temp_comp_lists_cle <- list()
est_list_cle <- list()
M = 5
for(m in 1:M){

temp_comp_lists_cle[[m]] <- complete_stick_list[[m]][-c(1:367),]
temp_comp_lists_cle[[m]]$time <- times_vector


est_list_cle[[m]] <- temp_comp_lists_cle[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(cle), se = sqrt(var(cle)/n()), n())
est_list_cle[[m]]$imp = m
}

est_list_cle

combined_est_cle <- do.call(rbind,est_list_cle)


#Calculating error bars of combined estimates

combined_est_errors_cle <- combined_est_cle %>% group_by(gender, time) %>% summarize(x_bar_bar  = mean(xbar), B_m = var(xbar), W_m = mean(se^2), se_xbar_bar = sqrt(W_m+(1+1/5)*(B_m)), r = (1+1/5)*B_m/W_m, U = (5-1)*(1+(1/r)^2), t_val = qt(0.05/2, df = U, lower.tail = FALSE), uci = x_bar_bar + t_val * se_xbar_bar, lci = x_bar_bar - t_val * se_xbar_bar, t_se = t_val * se_xbar_bar)

combined_est_errors_cle %>% select(gender, time,x_bar_bar, lci, uci)


ggplot(data= combined_est_errors_cle, aes(x = time, y = x_bar_bar, col = gender))+geom_point()+geom_line() + geom_errorbar(aes(ymin = lci, ymax = uci),
     width = 0.2) + labs(x = "Time", y = "Mean of imputed mean CLE")

```

```{r}
#Using MICE imputation to get imputed datasets (function)

impute_stickledata <- function(M, dataset){
  imputed_stick_ <- mice(dataset,m=M)
  complete_stick_list <- list()
  for(m in 1:M){
    complete_stick_list[[m]] <- complete(imputed_stick_, m)
  }
  return(complete_stick_list)
}

```
```{r}
# My debugging process. Found out about enquo() and !!, which allows for inputting variables within a function.
# pretty, pretty cool

# test_ <- function(M,variable,data){
#   variable <- enquo(variable)
#   temp_comp_lists <- list()
#   est_list <- list()
#   for(m in 1:M){
#   temp_comp_lists[[m]] <- data[[m]][-c(1:367),]
#   temp_comp_lists[[m]]$time <- times_vector
#   
#   
#   est_list[[m]] <- temp_comp_lists[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(!! variable), se = sqrt(var(!! variable)/n()), n())
#   est_list[[m]]$imp = m
#   }
#   return(est_list)
# }
# 
# testing_ <- impute_stickledata(5,stick_whole)
# test_(5, stl, testing_)


```


```{r}
completed_stick_lists <- impute_stickledata(5,stick_whole)
#doing the grouping and summarizing for all imputed datasets function
imputed_means <- function(M,target_variable){
target_variable <- enquo(target_variable)

temp_comp_lists <- list()
est_list <- list()

# completed_stick_lists <- impute_stickledata(M,stick_whole). I could leave this in here but don't necessarily need it
for(m in 1:M){

temp_comp_lists[[m]] <- completed_stick_lists[[m]][-c(1:367),]
temp_comp_lists[[m]]$time <- times_vector


est_list[[m]] <- temp_comp_lists[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(!! target_variable), se = sqrt(var(!! target_variable)/n()), n())
est_list[[m]]$imp = m
}


# stacking the datasets
combined_est <- do.call(rbind,est_list)
return(combined_est)
}

```

```{r}
#Function for combined SE
combined_se_function <- function(M, variable){
variable <- enquo(variable)
combined_estimate <- imputed_means(5, !! variable)
  
combined_est_errors <- combined_estimate %>% group_by(gender, time) %>% summarize(x_bar_bar  = mean(xbar), B_m = var(xbar), W_m = mean(se^2), se_xbar_bar = sqrt(W_m+(1+1/5)*(B_m)), r = (1+1/5)*B_m/W_m, U = (5-1)*(1+(1/r)^2), t_val = qt(0.05/2, df = U, lower.tail = FALSE), uci = x_bar_bar + t_val * se_xbar_bar, lci = x_bar_bar - t_val * se_xbar_bar, t_se = t_val * se_xbar_bar)

return(combined_est_errors)
}

combined_se_function(5, stl)

```

```{r}
#function for ggplot of desired var
stickle_imputation_plotting <- function(M, var){
  v <- as.character(substitute(var))
  var =enquo(var)
  string = paste("mean of ", v)
  combined_errors <- combined_se_function(5, !! var)
  
  bb<- ggplot(data= combined_errors, aes(x = time, y = x_bar_bar, col = gender))+geom_point()+geom_line() +   geom_errorbar(aes(ymin = lci, ymax = uci, group = gender),
     width = 0.2)+ labs(x = "Time", y = string)

  return(bb)
}
stickle_imputation_plotting(5,stl)


```

```{r}



```





